Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-31 Manuscript/manuscript.Rmd
intro edits

A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, class imbalance was not addressed and instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any transcription factors, resulting in a zero inflated model. This, likely being further exacerbated by the class imbalance. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS). However, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. There was also a consensus on feature engineering where both methods used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions.

<!--
Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites (TFBS) 
on GM12878 [@mourad2016computational]. High resolution Hi-C data at 1 kb was used and percentage of overlap between regions defining TFBS and 1 kb bins was used for feature engineering. However, class imbalance was not adressed and, instead area under the receiver operating curve (AUROC) was used to evaluate model performance. This is problematic since it has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications in an L2-based regularized linear model [@hong2017computational]. Both used re-sampling techniques prior to modeling, with both implementing random under-sampling to balance TAD and non-TAD boundary regions. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Furthermore, there was not a consensus in either resolution or feature engineering. The BART model was implemented on 5 kb Hi-C data using sequence read count between ChIP-seq data and 5 kb bins, while the L2-based regularized linear model was implemented on filtered TADs greater than 200 kb in length using the elemental count that appeared in each 150 kb bin. 
-->

To date, there has yet to be established a unified approach toward TAD boundary prediction. Therefore, in this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller, these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome

>>>>

A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, class imbalance was not addressed and instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. This, likely being further exacerbated by the class imbalance. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in basepairs between ChIP-seq annotations and genomic bins in TAD boundary prediction.

Therefore, given the many pitfalls encountered by previous methods we set out to establish a unified approach toward TAD boundary prediction. In this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances for TAD boundary prediction. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller, these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome.
